---
title: "Explorating ggplot2"
author: "Santina"
date: "Monday, September 29, 2014"
output: html_document
---

In this R markdown I'm going to try out some ggplot2 features I just learned from Prof. Jenny Bryan's [tutorial](https://github.com/jennybc/ggplot2-tutorial) on the [gapminder data](http://www.stat.ubc.ca/~jenny/notOcto/STAT545A/examples/gapminder/data/gapminderDataFiveYear.txt) like in the last Rmd assignment. 

First let's see what variables we can play with: 

```{r}
data  <- read.delim("gapminderDataFiveYear.txt")
str(data)
```

So we have country, continent, pop (population), gdpPercap, lifeExp, and year. That's plenty to work with. Since lifeExp has been explored quite throroughly in the tutorial, I'll focus on other variables. 

#Population Boom 
I'm interested to find out the trend of the population growth globally and be able to visualize which groups of countries (i.e continents) contribute to more/less of the growth and to what extend. 

So we need to break down the problem step by step.  
- sum population of all countries by year 
- construct a smooth histogram 
- use color to distinguish continents 

Step 1: 
```{r, echo=FALSE}
plot(cars)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
