---
title: "HW8: Cleaning data"
author: "Santina"
date: "Thursday, November 6 2014"
output: 
  html_document:
    keep_md: true
    toc: yes
    toc_depth: 4 
---

In this assignment, we are going to attemp to clean up the gapminder data that has been purposely messed up for the purpose of this assignment. We would use some techniques that have to do with manipulating data/table and regular expressions. 

First get some useful packages loaded first 

```{r, message=FALSE}
library(stringr) # for regular expression
library(plyr)       # for easy computation with data frames
library(dplyr)      # do this after loading plyr
library(knitr)      # for rendering pretty tables
library(grid)
library(gridExtra)  # arranging graph
```

# load the dirty gapminder data 
```{r, results='asis'}
# read our dirty gapminder file 
dirty_file <- read.delim("gapminderDataFiveYear_dirty.txt")
kable(head(dirty_file))
str(dirty_file)
```

We can see that there are 151 different regions, each is the country name and continent combined together. 

## strip.white argument 

Let's examine what strip.white argument does to a data.frame in reading a file. 

```{r, results='asis'}
dirty_file2 <- read.delim("gapminderDataFiveYear_dirty.txt", strip.white = TRUE)
kable(head(dirty_file2))
str(dirty_file2)

```

I don't see much difference, exacept that the number of region factor levels has changed. With `strip.white = FALSE` by default , it's 151, and now it's 148. Let's examine it closer: 

```{r}

```