---
title: "HW12: Data from the Web"
author: "Santina Lin"
date: "Friday, December 05, 2014"
output:
  html_document:
    keep_md: true
    toc: yes
    toc_depth: 4 
---

# Goal
From our last week of STAT545, we learned about the amazing packages on ROpenSci through [in-class-demo](http://stat545-ubc.github.io/webdata02_activity.html) and how to [get data from the web](http://stat545-ubc.github.io/webdata03_activity.html) by reading the website. 

This is the [assignment](http://stat545-ubc.github.io/hw12_data-from-web). I will focus on : 
- combining two different datasets together
- comining different packages to make new things
- attempt to scrap data 


First let's load the housekeeping stuff: 
```{r, message=FALSE}
# ==== Packages ====
library(ggplot2)    # for making plots
library(ggthemes)   # for customizaing ggplot graphs 
library(scales)     # for graphs scale
library(plyr)       # for easy computation with data frames
library(dplyr)      # do this after loading plyr
library(knitr)      # for rendering pretty tables
```

# Combine Gapminder and Geonames

First load the essentials for this section 

```{r, message=FALSE}
# ==== Geonmaes package ====
#devtools::install_github("ropensci/geonames") 
options(geonamesUsername = "santina")
library(geonames)

# ==== Gapminder data ====
gdURL <- "http://tiny.cc/gapminder"
gapminder <- read.delim(file = gdURL) 
```

Now time to try answeirng Jenny's question: "What is the relationship between per-capita GDP and the proportion of the population which lives in urban centers?"
To answer this we need to know:
- the cities in each country, assuming urban population means people living in cities
- the population in those cities
- how many people live in the cities centers, in each country

## get a list of countries to investigate
To see a trend, we need that information for each country. I suspect that gapminder might have less data. Let's check. 
```{r}
nlevels(gapminder$country)
countryInfo <- GNcountryInfo()
summary(countryInfo)
```
Indeed we see that gapminder has less countries. Also all the entry in countryInfo are characters.

Since we need information from both gapminder and the information from `geonames` packages, we'll use the countries with information in both. 

```{r}
common_countries <- intersect(gapminder$country, countryInfo$countryName)
length(common_countries)
```
There are only `r length(common_countries)` of them! What are they? 
```{r}
common_countries
```
So these are the countries we'll look at. It might have been due to spelling or spacing that some countries were picked out. 

## Find urban population in these countries 

We'll write a function to find the total urban population in these countries. 
```{r}

urbanPopulation <- adply(common_countries, .margins = 1, function(country){
  # filter the countryInfo by the country
  sub <- countryInfo %>%
    filter(countryName == country)
  # use the information of coordinates to get cities in that area
  cities <- GNcities(north = sub$north, east = sub$east, 
                     south = sub$south, west = sub$west)
  # further filter the cities to make sure they belong to the country we're looking at
  # notice "countrycode" vs "countryCode"
  cities <- cities %>% filter(countrycode == sub$countryCode) 
  
  #sum the population and store the information in a list to be bind into a dataframe
  totalPopulation <- sum(as.numeric(cities$population))
  result <- c(country, totalPopulation)
  
}) 
head(urbanPopulation)

```
Oh guess what, I wrote that `adply()` from scratch in one shot and code compiled on the first try. Yeah look at my R programming muscles I got from lifting homework and peer reviews every week. 

Fixing up the data frame: 
```{r}
# take out the first column 
urbanPopulation <- urbanPopulation[, c(2,3)]
colnames(urbanPopulation) <- c("Country", "Urban_Population")
head(urbanPopulation)
```

##Combine gapminder and the urban data set

Now let's bind more information from gapminder to this data frame. 
```{r}
gapminder_sub <- gapminder %>% 
  subset(country %in% common_countries) %>%
  subset(year == 2007)
  
urbanPopulation <- cbind(urbanPopulation, gapminder_sub$pop, gapminder_sub$gdpPercap)

# fix up the column names 
colnames(urbanPopulation) <- c("Country", "Urban_Population", "Total_Population", "GDP")
summary(urbanPopulation)
```
The field for urban population is seemed as charcaters so we'll change that. 

```{r}
urbanPopulation$Urban_Population  <- as.numeric(urbanPopulation$Urban_Population)
summary(urbanPopulation)
head(urbanPopulation)
```

##Calculate the urban ratio and its relationship with GDP

Now we just need an extra column for the urban population raio 
```{r}
ratios <- ddply(urbanPopulation, ~Country, function(data){
  ratio <- (data$Urban_Population)/(data$Total_Population)
})

head(ratios)
colnames(ratios) <- c("Country", "Urban_Ratio")
urbanPopulation_Ratio_GDP <- cbind(urbanPopulation, "Urban_Ratio" = ratios$Urban_Ratio)
```

Now we can graph to see if there's any relationship

```{r}
ggplot(urbanPopulation_Ratio_GDP, aes(x=Urban_Ratio, y=GDP)) + 
  geom_point(size=3) +
  ggtitle("Urban Population Ratio vs GDP") + 
  theme_economist()
```
Haha doesn't seem very correlated at all. There are some factors that have an effect on the analysis, such as how we define "urban" in this analysis, the information regarding to urban population (we don't know which year it is from), etc. 


# Combine Rplos and rebirds 
Now I'm interesting in knowing about the number of publications on birds. To do that, the easiest way is to get a list of birds and then use `rplos::plosword`, which allow you to search multiple terms at once and return the number of publications for each, with the options to view that result in a histogram. 

Pacakges involved: [rbirds](https://github.com/ropensci/rebird) and [rplos](https://github.com/ropensci/rplos)

Let's try searching for birds in Canada. 
```{r}
library(rplos)
library(rebird)
birds_Canada <- ebirdregion(region = 'CA')
head(birds_Canada)
length(unique(birds_Canada$sciName))
```
There are so many types of birds!  Maybe if I study one bird each day for a year, then I'll get to know all the bird species in Canada. Okay back to the topic: 

Let's search the number of publications

```{r}
# need PlosApiKey, 
# out <- plosword(birds_Canada$sciName)
```

# Aiming higher: combining rMaps and rWBclimate 
# Leaf in the wind: Grab data from Cystic Fibrosis mutation database 
