
Get data from the web 

```{r}
library(rebird)
library(knitr)
```
It's for a ebird data 

```{r}
ebirdgeo(species = "spinus tristis", lat = 42, lng = -76)
# given scientific name and the location 
ebirdgeo(species = "poecille gambell", lat = 42, lng = -123)
# it knows where you are too
ebirdgeo(species = 'buteo lagopus')
```

```{r results ='asis'}
vanbirds <- ebirdgeo(lat = 49.25, lng = -123.1)
kable(head(vanbirds)) # a bird list for vancouver 
```

```{r}
ebirdregion("US-IL") # given a region code, search the website 
```

# search geographic info : 'geonames'
```{r}
install.packages("rjson")
# downloading from github so to make sure it's most up-to-date 
devtools::install_github("ropensci/geonames") 
library(geonames)
```

# API and authentification
We're gonna use geonames to get geo info (? not sure)

1. geonames.org/login, make an account 
2. http://www.geonames.org/enablefreewebservice
3. go to this site, which enable you to get data in a more programmer way 

```{r}
options(geonamesUsername = "santina")
```

## let's get all countries in the world
```{r}
countryInfo <- GNcountryInfo()
head(countryInfo)

library(dplyr)
glimpse(countryInfo) # structure but easier to read 
```

Use `?functionName()` to see the source (url ) of the info

```{r}
sub <- countryInfo %>%
	filter(countryName == "France")

gcities <- GNcities(north = sub$north, 
										east = sub$east, 
										south = sub$south, 
										west = sub$west, 
										maxRows = 500)
head(gcities)
```

## how many birds have been seen in France? 

```{r}
frenchbirds <- countryInfo %>%
	filter(countryName %in% c("France", "Canada")) %>% # tried Germany, didn't work
	group_by(countryName) %>% # do by countryName (each country)
	do(allbirds = ebirdregion(.$countryCode)) # sticks all info into a cell of the datamframe


class(frenchbirds[[1,2]])
head(frenchbirds[[1,2]])
```

# by wikipedia
(missed)
```{r}

```


# use PLOS API 

```{r }
install.packages("rplos")
library("rplos") #http://ropensci.org/tutorials/rplos_tutorial.html
```

Get this account 
- http://alm.plos.org/
- save your API key as an environmental variable for privacy

```{r}

```