---
title: "HW5: Be the boss of my factors"
author: "Santina"
date: "Thursday, October 16, 2014"
output: 
  html_document:
    keep_md: true
    toc: yes
    toc_depth: 4 
---

# Introduction 
In this assignment, we are going to see how to be in [control of factors](http://stat545-ubc.github.io/block014_factors.html) in a data set, using the provided [Gapminder excerpt](http://www.stat.ubc.ca/~jenny/notOcto/STAT545A/examples/gapminder/data/gapminderDataFiveYear.txt) as an example, and experiment with [reading and writing files](https://github.com/STAT545-UBC/STAT545-UBC.github.io/blob/master/cm011_files-out-in-script.r). 

It will follow roughly (with perhaps some experimentations on the side) the assignment guideline. This is the last assignment of STAT545! I can't wait for the second half of the course, which will be even more exciting! 

As always, let's get the gapminder data. Since it's from an URL, I am going to try to download it by reading it into R and saving it as a text file in my current directory. 

## Notes to myself 
Just some notes to remind me what I can use 
- writing to and reading from files
  * use `write.table()` and `read.delim()` or `read.table()`, make sure to experiment with different arguments 
- writing and reading R objects
  * RDS format: `saveRDS()`, `readRDS()`
  * plain text format (often preferable):`dput()` and `dget()` 

## same old stuff: packages and the gapminder data
load some packages 
```{r, message=FALSE}
library(ggplot2) # for making plots
library(ggthemes)# for customizaing ggplot graphs 
library(scales)  # for graphs scale
library(plyr)    # for easy computation with data frames
library(dplyr)   # do this after loading plyr
library(knitr)   # for rendering pretty tables
```

Loading our gapminder data 
```{r}
gpURL <- "http://www.stat.ubc.ca/~jenny/notOcto/STAT545A/examples/gapminder/data/gapminderDataFiveYear.txt"
dataExcerpt <- read.delim(file = gpURL) 
str(dataExcerpt)
#saving it as "gapminder_excerpt.txt" in my current homework directory
write.table(dataExcerpt, "gapminder_excerpt.txt") 
```

Let's try to see how well we can read from the text file we just created. 
```{r}
gapExcerpt <-  read.delim(file = "gapminder_excerpt.txt")
str(gapExcerpt)

```
Hum.... a little massier than the result of `str(dataExcerpt)`. Let's come back to this later. 

# Drop Oceania 

```{r}


```


